---
title: Gene set searches with the Gesel database
author:
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com
package: gesel
date: "Revised: November 4, 2024"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Working with the Gesel database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
library(BiocStyle)
self <- Githubpkg("gesel-inc/gesel-R", "gesel")
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
```

# Introduction

The `r self` package implements an R client to the [Gesel database](https://doi.org/10.21105/joss.05777) for gene set searching.
The Gesel database is a collection of static files containing information about gene sets, which can be hosted anywhere - via a file server, on a HPC's shared filesystem etc. 
Clients like the `r self` package then use HTTP range requests to perform a variety of queries.
No custom server logic is required, greatly reducing effort and cost required to keep Gesel up and running.
Most database files do not need to be downloaded to the client, allowing us to easily scale with increasing numbers of gene sets in the database.

# Finding overlaps

The raison d'etre of the Gesel database is to find overlaps between known gene sets and a user-supplied list of genes.
Let's say we have several human genes of interest:

```{r}
my.genes <- c("SNAP25", "NEUROD6", "GAD1", "GAD2", "RELN")
```

We map these to Gesel gene indices, which are Gesel's internal identifiers for each gene.

```{r}
library(gesel)
gene.idx <- searchGenes("9606", my.genes)
gene.idx # this is a list of integer vectors, in case of 1:many mappings.
gene.idx <- unlist(gene.idx)
fetchAllGenes("9606")[gene.idx,] # double-checking that we got it right.
```

Now we find all sets with overlapping genes.
The `findOverlappingSets()` identifies all sets with one or more overlaps with the user-suppleid set.

```{r}
overlaps <- findOverlappingSets("9606", gene.idx, counts.only=FALSE)
head(overlaps$overlap) # set index and the identities of overlapping genes.
```

The set index can be converted back to information about each set, along with the collection from which it came:

```{r}
set.info <- fetchSomeSets("9606", head(overlaps$overlap$set)) # getting details of the top sets
set.info
col.info <- fetchSomeCollections("9606", set.info$collection) # getting details of the collections
col.info
```

The various statistics produced by `r self` can be also used to perform a simple hypergeometric test for enrichment.
This often yields a more relevant ranking than the absolute number of overlaps as the hypergeometric p-value considers the size of the gene sets.

```{r}
set.sizes <- fetchSetSizes("9606")[overlaps$overlap$set]
p <- phyper(
    # Subtract 1 to account for probability mass at the observed number of
    # overlaps when computing the upper tail.
    q=lengths(overlaps$overlap$genes) - 1L,

    # Number of genes in the gene set.
    m=set.sizes,

    # Number of genes not in the gene set. We use effectiveNumberOfGenes() to
    # ignore irrelevant pseudogenes, predicted genes, etc. in the database.
    n=effectiveNumberOfGenes("9606") - set.sizes,

    # Number of genes in the user-supplied list of genes that are present in
    # at least one Gesel set (ignore typos and other unknown genes).
    k=overlaps$present,

    lower.tail=FALSE
)
head(p)
```

# Searching on text

We can also search for gene sets based on the text in their names or descriptions.
The `searchSetText()` function accepts a query string with multiple words and the usual `*` (any characters) and `?` (one character) wildcards.

```{r}
chits <- searchSetText("9606", "cancer")
fetchSomeSets("9606", head(chits))
ihits <- searchSetText("9606", "innate immun*")
fetchSomeSets("9606", head(ihits))
thits <- searchSetText("9606", "cd? t cell")
fetchSomeSets("9606", head(thits))
```

Users can construct powerful queries by intersecting the sets recovered from `searchSetText()` with those from `findOverlappingSets()`.

```{r}
cancer.sets <- intersect(chits, overlaps$overlap$set)
info <- fetchSomeSets("9606", cancer.sets)
m <- match(cancer.sets, overlaps$overlap$set)
info$count <- lengths(overlaps$overlap$genes)[m]
info$pvalue <- p[m]
info[head(order(info$pvalue)),]
```

# Fetching all data

Gesel is designed around partial extraction from the database files,
but it may be more efficient to pull all of the data into the R session at once.
This is most useful for the gene set details, which can be retrieved _en masse_:

```{r}
set.info <- fetchAllSets("9606")
nrow(set.info)
head(set.info)
```

The set indices can then be used to directly subset the `set.info` data frame by row.
In fact, calling `fetchSomeSets()` after `fetchAllSets()` will automatically use the data frame created by the latter,
instead of attempting another request to the database.

```{r}
head(set.info[cancer.sets,])
head(fetchSomeSets("9606", cancer.sets)) # same results
```

The same approach can be used to extract collection information, via `fetchAllCollections()`;
gene set membership, via `fetchGenesForAllSets()`;
and the sets containing each gene, via `fetchSetsForAllGenes()`.

# Advanced use 

`r self` uses a lot of in-memory caching to reduce the number of requests to the database files within a single R session.
On rare occasions, the cache may become outdated, e.g., if the database files are updated while an R session is running.
Users can prompt `r self` to re-acquire all data by flusing the cache:

```{r}
flushMemoryCache()
```

Applications can specify their own functions for obtaining files (or byte ranges thereof) in each `r self` function.
For example, on a shared HPC filesystem, we could point `r self` towards a directory of Gesel database files.
This provides a performant alternative to HTTP requests for an institutional collection of gene sets.

# Session information {-}

```{r}
sessionInfo()
```
